Modal Verification Patterns for Systems

### 2025-06-23 paper.tex
started working on comments

%\input{chapters/squeezing.tex}
\begin{abstract}
 Although they differ in the functionality they offer, low-level systems exhibit certain patterns of design and utilization of computing resources.

  In this paper, we argue the position that \emph{modalities}, in the sense of modal logic, 
  should be a go-to approach when specifying and verifying low-level systems code.
  We explain how the concept of a \emph{resource context} helps guide the design of new modalities
  for verification of systems code, and
  we justify our perspective by discussing prior systems that have used modalities for systems verification
  successfully, arguing that they fit into the verification design pattern we articulate,
  and explaining how this approach might apply to other systems verification challenges.

  \todo[inline]{Colin says: how about this framing instead of the second paragraph above:}
  \begin{comment}
  In this paper we examine how modalities have emerged as a common structure in formal verification
  of low-level software, and explain how many recent examples naturally share 
  common structure in the relationship between the modalities and software features they are used
  to reason about.
  We explain how the concept of a \emph{resource context} (a class of system resources
  to reason about) naturally corresponds to families of modal operators indexed by
  system data, and how this naturally leads to using modal assertions to describe
  \emph{resource elements} (data in the relevant context).
  We also describe extensions of this idea to ongoing work in formal verification
  of filesystems code.
  \end{comment}
  \todo[inline,color=cyan]{Ismail, why did you comment out my suggestion and keep the original?}
\end{abstract}

@@

\paragraph{Hybrid Logic} Hybrid logics~\cite{blackburn1995hybrid,goranko1996hierarchies,gargov1993modal,areces2001hybrid,brauner2010hybrid} are a class of modal logics that add two primary new concepts to a logic. Nominals $\l \in \textsf{Loc}$ uniquely identify points in a model (e.g., a particular state), so there is exactly one point in the model where a nominal $\loc$ is true. Satisfaction operators are modal operators indexed by nominals, which enable claims about the truth of another proposition at some arbitrary point in the model identified by a nominal. Traditionally $@_l\varphi$ asserts that $\varphi$ is true in the state identified by $l$. Conceptually, given our context, we might like $l$ to indicate truth in a particular data structure $l$.  However, most prior work on hybrid logics works in classical logics rather than a substructural setting like separation logic. 
We adapt these ideas to assertions of the form $@[l]\varphi$ asserts that $\varphi$ is true in a data structure with \emph{logical} state $\sigma$ \emph{at location} $l$.\todo{seems out of place?}
In practice we must design variants of this for each kind of structure we wish to apply this to, e.g. tree nodes, free list, or snapshots.
%This section provides details of and models for each modal abstraction, focusing on how each data structure can act as a model for a data-structure-specific logic.
%We continue to motivate the development with aspects of the write amplification process described in the previous section; Section \ref{sec:evaluation} provides further examples using the modal abstractions in proofs.
\paragraph{Temporal Operators}
The best known class of modal logics in the systems community is undoubtedly temporal logics (whether \textsc{LTL}~\cite{pnueli1977temporal}, \textsc{CTL}~\cite{emerson1982using}, \textsc{TLA}~\cite{lamport1994temporal}, or others),
due in part to Lamport's influence~\cite{lamport2002specifying} leading to its not-infrequent use in specifying
distributed algorithms \cite{ongaro2014search,ailijiang2019wpaxos}.
\paragraph{Nominalization}
Some classes of assertions benefit specifically from \emph{naming} the explicit conditions where they are true (as opposed to simply requiring them to be true \emph{somewhere} or \emph{everywhere} as in the most classic modal operators).
This naming generally resembles
the \emph{satisfaction} operator of 
\emph{hybrid logic}~\cite{blackburn1995hybrid,areces2001hybrid}: $@_\iota P$ which evaluates the truth of $P$ at the named (Kripke model) state $\iota$. For this reason we refer to the general idea of naming circumstances explicitly as \emph{nominalization},
even though the examples we discuss are not necessarily actually hybrid logics.

@@

\todo[inline,color=cyan]{Ismail, is this text copy-pasted from the VMM paper? 
It has phrasing like ``In this paper...''
That's actually a major issue, self-plagiarism of another peer-reviewed piece of work can get
you into a lot of trouble.
}
On the x86-64 architecture, the input to the \textsc{MMU}'s address translation is a sparse hierarchical set of tables called \emph{page tables} (referring to pages of memory). As Figure \ref{fig:pagetables} shows, address translation proceeds by repeatedly taking designated slices of the virtual address and indexing into a table. On x86-64, standard configurations use 4 levels of page tables, labelled levels 4 through 1, with lookups in the level 1 page table resulting in the actual page of physical memory holding the requested data, and the low-order 12 bits being used to index into this page.\footnote{Technically levels 1--3 have designated names originating when there were fewer levels. We omit those names to avoid confusion, and for greater consistency with the recent addition of a 5th level on the most recent processors, where levels 4 and 5 are simply known by their numbers.}  Because address translation traverses a series of tables, the translation process or algorithm is sometimes referred to as a \emph{page-table walk}. While Figure \ref{fig:pagetables} and most of our details are specific to the x86-64 architecture, ARMv8 (a.k.a.\ \texttt{aarch64}), RISC-V, and PowerPC use similar hierarchical page tables for address translation.

>>>>

%\input{chapters/squeezing.tex}
\begin{abstract}
 Although they differ in the functionality they offer, low-level systems exhibit certain patterns of design and utilization of computing resources.
\begin{comment}
  In this paper, we argue the position that \emph{modalities}, in the sense of modal logic, 
  should be a go-to approach when specifying and verifying low-level systems code.
  We explain how the concept of a \emph{resource context} helps guide the design of new modalities
  for verification of systems code, and
  we justify our perspective by discussing prior systems that have used modalities for systems verification
  successfully, arguing that they fit into the verification design pattern we articulate,
  and explaining how this approach might apply to other systems verification challenges.
\end{comment}
  \todo[inline]{Colin says: how about this framing instead of the second paragraph above:}

  In this paper we examine how modalities have emerged as a common structure in formal verification
  of low-level software, and explain how many recent examples naturally share 
  common structure in the relationship between the modalities and software features they are used
  to reason about.
  We explain how the concept of a \emph{resource context} (a class of system resources
  to reason about) naturally corresponds to families of modal operators indexed by
  system data, and how this naturally leads to using modal assertions to describe
  \emph{resource elements} (data in the relevant context).
  We also describe extensions of this idea to ongoing work in formal verification
  of filesystems code.
  \todo[inline,color=cyan]{Ismail, why did you comment out my suggestion and keep the original?}
  \todo[inline,color=pink]{Sorry, I thought I did the other way}
\end{abstract}

@@

\paragraph{Hybrid Logic} Hybrid logics~\cite{blackburn1995hybrid,goranko1996hierarchies,gargov1993modal,areces2001hybrid,brauner2010hybrid} are a class of modal logics that add two primary new concepts to a logic. Nominals $\l \in \textsf{Loc}$ uniquely identify points in a model (e.g., a particular state), so there is exactly one point in the model where a nominal $\loc$ is true. Satisfaction operators are modal operators indexed by nominals, which enable claims about the truth of another proposition at some arbitrary point in the model identified by a nominal. Traditionally $@_l\varphi$ asserts that $\varphi$ is true in the state identified by $l$. Conceptually, given our context, we might like $l$ to indicate truth in a particular data structure $l$.  However, most prior work on hybrid logics works in classical logics rather than a substructural setting like separation logic. 
%We adapt these ideas to assertions of the form $@[l]\varphi$ asserts that $\varphi$ is true in a data structure with \emph{logical} state $\sigma$ \emph{at location} $l$.
\todo{seems out of place?} \todo[inline,color=pink]{addressed}
In practice we must design variants of this for each kind of structure we wish to apply this to, e.g. tree nodes, free list, or snapshots.
%This section provides details of and models for each modal abstraction, focusing on how each data structure can act as a model for a data-structure-specific logic.
%We continue to motivate the development with aspects of the write amplification process described in the previous section; Section \ref{sec:evaluation} provides further examples using the modal abstractions in proofs.
\paragraph{Temporal Operators}
The best known class of modal logics in the systems community is undoubtedly temporal logics (whether \textsc{LTL}~\cite{pnueli1977temporal}, \textsc{CTL}~\cite{emerson1982using}, \textsc{TLA}~\cite{lamport1994temporal}, or others),
due in part to Lamport's influence~\cite{lamport2002specifying} leading to its not-infrequent use in specifying
distributed algorithms \cite{ongaro2014search,ailijiang2019wpaxos}.
\paragraph{Nominalization}
Some classes of assertions benefit specifically from \emph{naming} the explicit conditions where they are true (as opposed to simply requiring them to be true \emph{somewhere} or \emph{everywhere} as in the most classic modal operators).
This naming generally resembles
the \emph{satisfaction} operator of 
\emph{hybrid logic}~\cite{blackburn1995hybrid,areces2001hybrid}: $@_\iota P$ which evaluates the truth of $P$ at the named (Kripke model) state $\iota$. For this reason we refer to the general idea of naming circumstances explicitly as \emph{nominalization},
even though the examples we discuss are not necessarily actually hybrid logics.

@@

\todo[inline,color=cyan]{Ismail, is this text copy-pasted from the VMM paper? 
It has phrasing like ``In this paper...''
That's actually a major issue, self-plagiarism of another peer-reviewed piece of work can get
you into a lot of trouble.
}\todo{inline,color=pink}{I am well-aware what counts as as (including self-plagiarism). These were annotated with todos including cautions for changing the text.}
On the x86-64 architecture, the input to the \textsc{MMU}'s address translation is a sparse hierarchical set of tables called \emph{page tables} (referring to pages of memory). As Figure \ref{fig:pagetables} shows, address translation proceeds by repeatedly taking designated slices of the virtual address and indexing into a table. On x86-64, standard configurations use 4 levels of page tables, labelled levels 4 through 1, with lookups in the level 1 page table resulting in the actual page of physical memory holding the requested data, and the low-order 12 bits being used to index into this page.\footnote{Technically levels 1--3 have designated names originating when there were fewer levels. We omit those names to avoid confusion, and for greater consistency with the recent addition of a 5th level on the most recent processors, where levels 4 and 5 are simply known by their numbers.}  Because address translation traverses a series of tables, the translation process or algorithm is sometimes referred to as a \emph{page-table walk}. While Figure \ref{fig:pagetables} and most of our details are specific to the x86-64 architecture, ARMv8 (a.k.a.\ \texttt{aarch64}), RISC-V, and PowerPC use similar hierarchical page tables for address translation.
